High-level cognition is supported by at least second order dynamic correlations in neural activity patterns

### 2019-09-02 paper/main.tex
working on results for synthetic data

Given a kernel function $\kappa_t(\cdot)$ for timepoint $t$,
evaluated at timepoints $\tau \in \left[ 1, ..., T \right]$, we
can update the static correlation formula in Equation~\ref{eqn:corr}
to estimate the \textit{instantaneous correlation} at timepoint $t$:
\begin{align}
  \mathrm{timecorr}_{\kappa_t}\left(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)\right) &= \frac{\sum_{\tau=1}^T \left( \mathbf{X}(\tau, i) -
                                       \widetilde{\mathbf{X}}_{\kappa_t}(i) \right)
                                 \left( \mathbf{X}(\tau, j) -
                                        \widetilde{\mathbf{X}}_{\kappa_t}(j)\right)}
              {\sqrt{\sum_{\tau=1}^T
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot,
                                                                                        i))
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, j))}},~\mathrm{where}\\\label{eqn:timecorr}
  \widetilde{\mathbf{X}}_{\kappa_t}(t, k) &= \sum_{\tau=1}^T
                       \kappa_t(\tau, k)\mathbf{X}(\tau, k),\\
  \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, k)) &= \sum_{\tau=1}^T
                                                  \left(
                                                  \mathbf{X}(\tau, k) -
                            \widetilde{\mathbf{X}}_{\kappa_t}(t, k) \right)^2.
\end{align}
Here $\mathrm{timecorr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j),
  \kappa_t)$ reflects the correlation at time $t$ between columns $i$
  and $j$ of $\mathbf{X}$, estimated using the kernel $\kappa_t$.

@@

\subsection*{Higher-order correlations}

@@

\section*{Results}
\textbf{JRM STOPPED HERE}

%overview: first validate our approach with synthetic data, then apply
%to fMRI data

@@

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/synthetic_data.pdf}
  \caption{\textbf{Dynamic correlation recovery with synthetic data. } Using synthetic data containing different
    underlying correlational structure,
   we test how well we can recover dynamic correlation matrices using
   different kernels when compared to groud truth.  We compare the results using a delta kernel
   with averaged results from several kernels (Gaussian, Laplace, and
   mexican hat) and several widths (5, 10, 20, and 50).  We plot recovery using of datasets containing the
    following underlying structure:
    \textbf{a. Constant. b. Random. c. Ramping.  d. Block.}}
  \label{fig:synthetic_data}
\end{figure}

>>>>

Given a kernel function $\kappa_t(\cdot)$ for timepoint $t$,
evaluated at timepoints $\tau \in \left[ 1, ..., T \right]$, we
can update the static correlation formula in Equation~\ref{eqn:corr}
to estimate the \textit{instantaneous correlation} at timepoint $t$:
\begin{align}
  \mathrm{timecorr}_{\kappa_t}\left(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)\right) &= \frac{\sum_{\tau=1}^T \left( \mathbf{X}(\tau, i) -
                                       \widetilde{\mathbf{X}}_{\kappa_t}(i) \right)
                                 \left( \mathbf{X}(\tau, j) -
                                        \widetilde{\mathbf{X}}_{\kappa_t}(j)\right)}
              {\sqrt{\sum_{\tau=1}^T
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot,
                                                                                        i))
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, j))}},~\mathrm{where}\\\label{eqn:timecorr}
  \widetilde{\mathbf{X}}_{\kappa_t}(t, k) &= \sum_{\tau=1}^T
                       \kappa_t(\tau, k)\mathbf{X}(\tau, k),\\
  \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, k)) &= \sum_{\tau=1}^T
                                                  \left(
                                                  \mathbf{X}(\tau, k) -
                            \widetilde{\mathbf{X}}_{\kappa_t}(t, k) \right)^2.
\end{align}
Here
$\mathrm{timecorr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j),
\kappa_t)$ reflects the correlation at time $t$ between columns $i$
and $j$ of $\mathbf{X}$, estimated using the kernel $\kappa_t$.  We
evaluate Equation~\ref{eqn:timecorr} in turn each pair of columns in
$\mathbf{X}$ and for kernels centered on each timepoint in the
timeseries, respectively, to obtain a $T$ by $K$ by $K$ timeseries of
dynamic correlations, $\mathbf{Y}$.  For convenience, we then reshape
the upper triangles of each timepoint's correlation matrix into a row
vector to obtain an equivalent $T$ by $\frac{K^2 - K}{2}$ matrix.

@@

\subsection*{Dynamic higher-order correlations}

@@

\section*{Results}
We sought to understand whether high-level cognition is supported by
dynamic patterns of high-order correlations.  To that end, we
developed a computational framework for estimating the dynamics of
high-order correlations in multivariate timeseries data (see
\textit{Dynamic inter-subject functional connectivity (DISFC)} and
\textit{Dynamic higher-order correlations}).  We evaluated the
efficacy of this framework at recovering known patterns in several
synthetic datasets (see \textit{Synthetic data}).  We then applied the
framework to a public fMRI dataset collected as participants listened
to an auditorally presented story, a temporally scrambled version of
the story, or underwent a resting state scan (see \textit{Functional
  neuroimaing data collected during story listening}).  We used the
relative decoding accuracies of classifiers trained on
different sets of neural features to estimate which types of features
reflected ongoing cognitive processing.

\subsection*{Recovering known dynamic correlations from synthetic data}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/synthetic_data.pdf}
  \caption{\textbf{Recovering known dynamic correlations from
      synthetic data.}  Each panel displays the average correlations between
    the vectorized upper triangles of the recovered correlation matrix
    at each timepoint and either the true underlying correlation at each timepoint or a
    reference correlation matrix.  (The averages are taken across 10
    different randomly generated synthetic datasets.) Error ribbons
    denote 95\% confidence intervals (taken across datasets). Different colors denote different
    kernel shapes, whereas the shading within each color type denotes
    kernel widths.  For a complete description of each synthetic
    dataset, see \textit{Synthetic data}.  \textbf{a. Constant
      correlations.} These datasets have a stable (unchanging)
    underlying correlation matrix.  \textbf{b. Random
      correlations.}  These datasets are generated using a new
    independently drawn correlation matrix at each new timepoint.  \textbf{c. Ramping
      correlations.}  These datasets are generated by smoothly varying
    the underlying correlations between the randomly drawn
    correlation matrices at first and last timepoints.  The left
    panel displays the correlations between the recovered dynamic
    correlations and the underlying ground truth correlations.  The
    middle panel compares the recovered correlations with the
    \textit{first} timepoint's correlation matrix.  The right panel
    compares the recovered correlations with the
    \textit{last} timepoint's correlation matrix.
    \textbf{d. Event-based correlations.}  These datasets are each
    generated using five randomly drawn correlation matrices that each
    remain stable for a fifth of the total timecourse.  The left panel
    displays the correlations between the recovered dynamic
    correlations and the underlying ground truth correlations.  The
    right panels compare the recovered correlations with the
    correlation matrices unique to each event.}
  \label{fig:synthetic_data}
\end{figure}

\textbf{JRM STOPPED HERE}

@@
